Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-12-14 Dail_Mad_ext.tex
Stochasticity section and random effect paragraph

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. Although demographic and environmental stochasticity
in $r$ could be modeled as functions of covariates or random effects, a
simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

\citet{dail_madsen:2011} suggested two distributions for
modeling
initial abundance: Poisson and negative binomial. We have
extended
their model to include another distribution, the zero-inflated
Poisson. This could be useful when, for example, one is modeling
the
abundance of several bird species with the same set of BBS
routespoint
count surveys for all species, but this set includes routes
sites
outside the range of some of the bird species. The distribution
of
initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

These models allow three sources of zero counts by observers: a
species was at a route but not detected; the route was within the
species' range but there were no birds at that site in that
year; and the route was outside the species' range. Furthermore,
detection, abundance, and zero-inflation can be modeled separately as
functions of different (or the same) covariates. For example, detection of
species x at site y in year z might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.  This approach combines elements of occupancy modeling
\citep{mackenzie_etal:2006} and abundance modeling.
%, and could prove useful
%for distinguishing species' fundamental and realized niches
%(Hutchinson 1957).

@@

\subsection{Environmental and demographic stochasticity}

Even though the DM model is conceptually very simple, even in
its most stripped down form it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
demographic stochasticity is often modeled as $r_t \sim
\text{Norm}(\mu_r, \sigma_r)$.

@@

We applied these models to North American Breeding Bird Survey
(BBS) data from 1966-2010 for two species in the bordering US states
Maryland and Virginia. For our focal species, we selected
ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
forest breeding migrant with a stable or increasing trend in the region (CITE)
and golden-winged warblers (\textit{Vermivora chrysoptera}), which
only breeds in the western parts of these states and has been declining in the
region (CITE).  %MORE INFORMATION ABOUT THE SPECIES?

@@

[paragraph on random effects]

@@

Fortunately, it is straight-forward to incorporate direct information about
detection probability and we recommend that this be done whenever
possible. The original method for proposed in the original paper
was to collected replicated counts within the primary periods during
a period in which the populations could be assumed to be closed. A
robust design could be used to combine multiple surveys per
primary period to increase the precision of the estimates. We envision
that multiple other options are available as well, such as removal,
time of detection, and distance modeling ([cites]). These could be
accomplished by extending the open population $N$-mixture model
in exactly the same way as the closed population version has been
extended \citep[e.g.,][]{royle_etal:2004}.

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually-interesting
extensions that we belive would be computationally prohibitive in many
cases. Nonetheless, we will discuss one---spatially-explict models of
immigration. 
In our models immigration is currently modeled as independent of
population sizes in other sites. Several alternatives suggest
themselves if we think about movement in a metapopulation
context ([cite]): immigration depending on total or mean
abundance across sites the previous time step; immigration
depending on abundance in nearby sites only; and immigration
depending on abundance at sites as a function of their distance
from the receiving site (LOTS OF CITATIONS). These models would
be based on the assumptions that either the study sites
spatially cover the whole range of the metapopulation or the
annual changes in the populations sampled are representative of
a larger metapopulation. Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework.

\begin{comment}

@@

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
classical and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\hl{[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]}

>>>>

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. Although demographic and environmental stochasticity
in $r$ could be modeled as functions of covariates or random effects, a
simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

\citet{dail_madsen:2011} suggested two distributions for
modeling
initial abundance: Poisson and negative binomial. We have
extended
their model to include another distribution, the zero-inflated
Poisson. This could be useful when, for example, one is modeling
the
abundance of several species with the same set of point
count surveys for all species, but this set includes routes
sites
outside the range of some of the species. The distribution
of
initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

These models allow three sources of zero counts by observers: a
species was at a site but not detected; the site was within the
species' range but there were no animals at that site in that
year; and the site was outside the species' range. Furthermore,
detection, abundance, and zero-inflation can be modeled separately as
functions of different (or the same) covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.  
%This approach combines elements of occupancy modeling
%\citep{mackenzie_etal:2006} and abundance modeling.
%, and could prove useful
%for distinguishing species' fundamental and realized niches
%(Hutchinson 1957).

@@

\subsection{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, even in
its most stripped down form it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\text{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models include demographic stochasticity (variation in population growth 
rate due to the randomness of birth and death processes) through random 
binomial draws for survival and Poisson draws for recruitment.  Our population
growth extensions use a single Poisson draw for abundance: $N_{it} \sim 
\text{Pois}(f(N_{it-1}))$.  This is also a commonly used (and mathematically equivalent) 
method of modeling demographic stochasticity 
\citep{bonsall_hastings:2004,melbourne_hastings:2008}.

One way to model environmental stochasticity (variation in population growth
rate due to stochastic environmental conditions) is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature and $K$ a function of rainfall:
\begin{gather}
r_{it} = \exp(r_{0} + r_{temp}T_{it}) \nonumber \\
K_{it} = \exp(K_{0} + K_{rain}R_{it})  \\
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
\label{eq:weather}
\end{gather}
where $T_{it}$ is site and year specific temperature, $R_{it}$ is site and year specific rainfall,
and the errors for the coefficients $r_{0}$, $r_{temp}$, $K_{0}$, and $K_{rain}$ are 
normally distributed.

Another way to model environmental stochasticity is as lognormal variation in observed local population 
growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
For example, for the Ricker-logistic + immigration model:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(\nu_{it} + r(1-N_{it-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
Three variations suggest themselves.  The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{it} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
In cases where sites can be close together but the range is broad, a multivariate normal distribution for $\nu_{it}$ with correlation coefficients
between sites in a year dependent on distance (geographic or otherwise) might be appropriate.

@@

We applied these models to North American Breeding Bird Survey
(BBS) data from 1966-2010 for two species in the bordering US states
Maryland and Virginia. For our focal species, we selected
ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}
and golden-winged warblers (\textit{Vermivora chrysoptera}), which
only breeds in the western parts of these states and has been declining in the
region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

@@

Even though the DM models and our extentions are conceptually simple, even in
their most stripped down form they include many random effects (the $N$'s, which
incorporate demographic stochasticity), making it challenging to compute the likelihood 
or implement MCMC algorithms. Nonetheless, in some cases additional random effects
may be of interest.  We have shown a few approaches for incorporating some of these
(environmental stochasticity and random observer effects).  Other approaches to modeling
each of these may also be worth trying.  Incorporating these random effects can increase
the difficulty of achieving model convergence, but they are certainly important.
[Something about how much including each of these random effects changes ovenbird estimates.]

@@

Fortunately, it is straight-forward to incorporate direct information about
detection probability and we recommend that this be done whenever
possible. The original method for proposed in the original paper
was to collected replicated counts within the primary periods during
a period in which the populations could be assumed to be closed. A
robust design could be used to combine multiple surveys per
primary period to increase the precision of the estimates. We envision
that multiple other options are available as well, such as removal,
multiple observer, and distance modeling \citep{williams_etal:2002}. These could be
accomplished by extending the open population $N$-mixture model
in exactly the same way as the closed population version has been
extended \citep[e.g.,][]{royle_etal:2004}.

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually-interesting
extensions that we belive would be computationally prohibitive in many
cases. Nonetheless, we will discuss one---spatially-explict models of
immigration. 
In our models immigration is currently modeled as independent of
population sizes in other sites. Several alternatives suggest
themselves if we think about movement in a metapopulation
context \citep{hanski:1998}: immigration depending on total or mean
abundance across sites the previous time step; immigration
depending on abundance in nearby sites only; and immigration
depending on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
be based on the assumptions that either the study sites
spatially cover the whole range of the metapopulation or the
annual changes in the populations sampled are representative of
a larger metapopulation. Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework.
% Could use a better reference than Hastings 1991 for the total abundance 
% immigration models, but haven't been able to find one.  And haven't found
% any references for nearby site immigration (maybe that's because it's not a great idea).
\begin{comment}

@@

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
classical and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}
